---
title: "DE for VBCS-721"
author: "Jay Kim & Daisy Fu"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    code_folding: hide
    self_contained: yes
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
---
# DE analysis 

JIRA ticket number: VBCS-721

Phage display data was obtained from the LiGA system. 

The following comparisons are performed.  
1. Anti_GS2 vs Control  
2. AntiLewisX vs Control  
3. CA19_9 vs Control  
4. ConA vs Control  
5. CTB vs Control  
6. ECL vs Control  
7. MAL_I vs Control  
8. Naive vs Control  
9. PNA vs Control  
10. TR4 vs Control  
11. VVL vs Control  
12. WFA vs Control  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=F, message=F, cache=F)
```

# Load packages

```{r load_libs}
rm(list=ls())
set.seed(0)

library(enrichplot)
library(tidyverse)
library(edgeR)
library(DESeq2)
library(stringr)
library(cowplot)
library(RColorBrewer)
library(ggrepel)
library(EnhancedVolcano)
# library(bbcRNA)
# library(clusterProfiler)
library(pcaExplorer)
library(ComplexHeatmap)
library(kableExtra)
# library(ggsci)
```

# Import counts and meta data


```{r load_data}
base <- "./dna_count_output_dir/"
output_dir <- 'de_out'
dir.create(output_dir)
col_name <- c("sdb_id", "seq_count")

## meta data load
meta_raw <- read_csv("BarcodeSampleMap.csv")
meta_df <- meta_raw %>% mutate(samp_file = paste0(Sample, "_", Replicate, ".txt"))
meta_df$sample <- paste0(meta_df$Sample, "_", meta_df$Replicate)
meta_df <- meta_df %>% dplyr::rename(Group = Sample) %>% dplyr::select(sample, Group, Replicate, Forward, Reverse, samp_file) %>% data.frame()
meta_df$sample <- str_replace(meta_df$sample, "-", "_")
meta_df$Group <- str_replace(meta_df$Group, "-", "_")
row.names(meta_df) <- meta_df$sample
meta_df$Replicate <- as.factor(meta_df$Replicate)
meta_df$Group <- as.factor(meta_df$Group)
meta_df$Group <- relevel(meta_df$Group, ref = "Control")
meta_df %>% dplyr::select(-sample) %>% DT::datatable()

file_list <- file.path(paste0(base, meta_df$samp_file))

## merge all dna count table files. use full_join to keep SDB IDs.
merged_df <- file_list %>% lapply(read_csv, col_names = col_name) %>% purrr::reduce(full_join, by = 'sdb_id')
merged_filt_df <- merged_df %>% dplyr::select(where(is.numeric))
file_name <- str_extract(file_list, "[:alnum:]*-?[:alnum:]*_[12]") %>% str_replace(.,"-", "_")

colnames(merged_filt_df) <- file_name
merged_filt_df <- merged_filt_df %>% add_column(sdb_id = merged_df$sdb_id, .before = 1)
# merged_filt_df %>% head(50) %>% view()

## replace NA with 0
merged_filt_df <- merged_filt_df %>% replace(is.na(.), 0)

## duplicated values found in dna column
count_df <- merged_filt_df %>% data.frame()
count_filt <- count_df
# count_filt <- count_df %>% distinct(dna, .keep_all = TRUE)
rownames(count_filt) <- count_filt$sdb_id
count_filt <- count_filt %>% dplyr::select(-sdb_id)
# write_csv(count_df, "2022.04.05.ver.1.merged_count_table.tsv")
```


# PCA 

DESeq2 is used to generate PCA here. 


```{r PCA, fig.height= 6.5, fig.width=8}

## now counts_24 and meta_df should be ready to feed into the dds or edgeR DGElist
## PCA using DESeq2

dds <- DESeqDataSetFromMatrix(countData = count_filt,
                              colData = meta_df,
                              design = ~Group)
## use vst
vsdata <- varianceStabilizingTransformation(dds, blind=FALSE)
dds <- DESeq(dds)

vsdata_colnames <- colnames(vsdata)

pcaData <- DESeq2::plotPCA(vsdata, intgroup=c("Group"), returnData=TRUE)

pcaobj_prop <- prcomp(t(assay(vsdata)))
pcaobj_samp <- prcomp(assay(vsdata))
percentVar <- round(pcaobj_prop$sdev ^2 / sum(pcaobj_prop$sdev ^ 2),3) * 100

# pc1_pc2_pca <- ggplot(pcaData, aes(PC1, PC2, color=Group)) +
#   geom_point(size=3) +
#   xlab(paste0("PC1: ",percentVar[1],"% variance")) +
#   ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
#   theme_bw() +
#   coord_fixed() + 
#   ggrepel::geom_text_repel(label = meta_df$sample, size = 4) + 
#   theme(legend.position = "none") #+ scale_color_ucscgb()
# 
# pc1_pc3_pca <- ggplot(pcaData, aes(PC1, PC3, color=Group)) +
#   geom_point(size=3) +
#   xlab(paste0("PC1: ",percentVar[1],"% variance")) +
#   ylab(paste0("PC3: ",percentVar[3],"% variance")) + 
#   theme_bw() +
#   coord_fixed() + 
#   ggrepel::geom_text_repel(label = meta_df$sample, size = 4) + 
#   theme(legend.position = "none")
# 
# pc2_pc3_pca <- ggplot(pcaData, aes(PC2, PC3, color=Group)) +
#   geom_point(size=3) +
#   xlab(paste0("PC1: ",percentVar[2],"% variance")) +
#   ylab(paste0("PC3: ",percentVar[3],"% variance")) + 
#   theme_bw() +
#   coord_fixed() + 
#   ggrepel::geom_text_repel(label = meta_df$sample, size = 4) + 
#   theme(legend.position = "none")

pca_pc1_pc2 <- pcaplot(vsdata, intgroup = "Group", pcX = 1, pcY = 2, title = "PC1 & PC2 - variance stabilizing transformed", ntop = Inf, text_labels = F)
pca_pc1_pc3 <- pcaplot(vsdata, intgroup = "Group", pcX = 1, pcY = 3, title = "PC1 & PC3 - variance stabilizing transformed", ntop = Inf, text_labels = F)
pca_pc2_pc3 <- pcaplot(vsdata, intgroup = "Group", pcX = 2, pcY = 3, title = "PC2 & PC3 - variance stabilizing transformed", ntop = Inf, text_labels = F)
pca_pc1_pc2
pca_pc1_pc3
pca_pc2_pc3

pcaobj_prop <- prcomp(t(assay(vsdata)))
pcascree(pcaobj_prop, type='pev', title="Proportion of explained of variance") + scale_y_continuous(labels= scales::percent)


```

# DE analysis 


```{r DE}
## check if the orders are right. 
all(rownames(meta_df) == colnames(count_filt)) 


y <- DGEList(count_filt, samples = meta_df, group = meta_df$Group)
design <- model.matrix(~0 + Group, data = meta_df)
# design <- model.matrix(~0 + Genotype + Batch, data = meta)
# intercept GenotypeKO and Batch1
rownames(design) <- colnames(y)


# keep <- rowSums(edgeR::cpm(y$counts) > 0.5) >= 2

## edgeR user guide page 13
keep <- filterByExpr(y)
table(keep) ## check how many genes will be kept and how many will be removed. 

y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

norm_counts <- edgeR::cpm(y, log=TRUE)

## export count data, which are normalized and CPM, note here, log transformation is log2 based. 
norm_counts_df <- norm_counts %>%
  as.data.frame() %>% 
  rownames_to_column(var = "sdb_id")
  
write_rds(norm_counts_df, paste0(output_dir, "/","norm_counts.rds"))
write_tsv(norm_counts_df, paste0(output_dir, "/","norm_counts.tsv"))
write_tsv(count_df, paste0(output_dir, "/", "raw_counts.tsv"))
write_rds(count_df, paste0(output_dir, "/","raw_counts.rds"))


y <- estimateDisp(y, design, robust=TRUE)
# plotBCV(y, main=paste0("BCV plot"))

fit <- glmQLFit(y, design, robust=TRUE)
# plotQLDisp(fit, main=paste0("QLDisp plot"))

# rm(contrasts_list)
 contrasts_list <- c(
                   "Anti_GM2-Control",
                   "AntiLewisX-Control",
                   "CA19_9-Control",
                   "ConA-Control",
                   "CTB-Control",
                   "ECL-Control",
                   "MAL_I-Control",
                   "Naive-Control",
                   "PNA-Control",
                   "TR4-Control",
                   "VVL-Control",
                   "WFA-Control"
                   )
 
qlf <- glmQLFTest(fit)
design <- model.matrix(~0+Group, data = y$samples)
colnames(design) <- gsub("Group", "", colnames(design))

list_with_names <- contrasts_list 
names(list_with_names) <- list_with_names

contrasts <- makeContrasts(contrasts=contrasts_list, levels=design)

qlf <- lapply(rlang::set_names(colnames(contrasts), colnames(contrasts)), 
               function(contrast){
                 glmQLFTest(fit, contrast=contrasts[,contrast])} )

res <- lapply(qlf, function(contrast) topTags(contrast, n = Inf))
save(y, qlf, fit, res, file = paste0(output_dir, "/","edgeR", "_data_file.RData"))

```

# Define function

```{r define_function}
DE_gene_by_lappy <- function(current_contrast) {
  qlf_current <- qlf[[current_contrast]]
  res_current <- res[[current_contrast]]
  toptags <- topTags(qlf_current, n = 10) %>% as.data.frame() #%>% dplyr::select(!c(Uniq_syms, entrez))
  cat(paste0("Results of comparison ", current_contrast), sep = "\n\n")
  cat(paste0("Samples and meta table for this comparison below: "), sep = "\n\n")
  current_samples <- meta_df[meta_df$Group %in% str_split_fixed(current_contrast, "-", 2)[1, ], ] ## 
  # print(current_samples %>% dplyr::select(sample, Group))
  
  cat(paste0(" "), sep = "\n\n")
  cat(paste0("Number of DE genes: "), sep = "\n\n")
  DE_sum <- summary(decideTests(qlf_current))
  print(DE_sum)
  
  cat(paste0(" "), sep = "\n\n")
  cat(paste0("Top 10 DE SDB ID sorted by FDR"), sep = "\n\n")
  print(toptags)
  
  # print(plotMD(qlf_current, hl.cex= 0.75, main = paste0("MA plot of ", current_contrast)))
  res_current$table <- rownames_to_column(res_current$table, var = 'sdb_id')
  print(EnhancedVolcano(res_current$table, 
                  lab = res_current$table$sdb_id,
                  x = 'logFC', pCutoff = 0.05, FCcutoff= 1,
                  y = 'FDR', subtitle = "",
                  ylab = "-log10(FDR)",
                  legendPosition = 'right',
                  legendLabels = c("Not Sig", expression(Log[2] ~ FC), "FDR", expression(FDR ~ and ~ log[2] ~ FC)),
                  caption="", title = paste0('Volcano plot of ', current_contrast)))

  ## plot heatmap
  heatmap_id <- qlf_current %>% topTags(., n = Inf) %>% as.data.frame() %>% rownames_to_column(var = 'sdb_id') %>% filter(abs(logFC) >= 1) %>% head(30)
  logCPM <- cpm(y, log = T)
  logCPM_df <- logCPM[, current_samples$sample]
  print(Heatmap(t(scale(t(logCPM_df))), 
          rect_gp = gpar(col = "white", lwd = 2), 
          row_names_gp = gpar(fontsize = 5), 
          row_dend_side = 'right',
          column_title = "Heatmap of the top 30 DE SDBs", 
          heatmap_legend_param = list(
                                title = "z-score")
          )
  ) # end print
  
  sams_wanted <- current_samples$sample
  
  raw_counts_wanted <- as.data.frame(fit$counts)  %>%
    .[, colnames(.) %in% sams_wanted] %>% rownames_to_column(., var = "sdb_id")
  
  norm_counts_wanted <- as.data.frame(norm_counts) %>%
    .[, colnames(.) %in% sams_wanted] %>% rownames_to_column(., var = "sdb_id")
  
  res_export <- left_join(res_current$table, norm_counts_wanted, by= "sdb_id") 
  res_export <- left_join(res_export, raw_counts_wanted, by = "sdb_id")
  
  colnames(res_export) <- colnames(res_export) %>% str_replace_all(., ".x", "_normalized_read") %>% str_replace_all(., ".y", "_raw_read")
  
  readr::write_tsv(res_export, file = paste0(output_dir,"/", "edgeR", "_", current_contrast, 
                                              "_DE_results.tsv"))
}

```

# DE gene results

## Anti_GM2 vs Control
```{r DE_comp1, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[1])

```

## AntiLewisX vs Control

```{r DE_comp2, fig.height=8, fig.width=9}

DE_gene_by_lappy(current_contrast = list_with_names[2])

```

## CA19_9 vs Control

```{r DE_comp3, fig.height=8, fig.width=9}

DE_gene_by_lappy(current_contrast = list_with_names[3])

```

## ConA vs Control

```{r DE_comp4, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[4])

```

## CTB vs Control

```{r DE_comp5, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[5])

```

## ECL vs Control

```{r DE_com6, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[6])
```

## MAL_I vs Control

```{r DE_comp7, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[7])
```

## Naive vs Control

```{r DE_comp8, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[8])
```

## PNA vs Control

```{r DE_comp9, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[9])
```

## TR4 vs Control

```{r DE_comp10, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[10])
```

## VVL vs Control

```{r DE_comp11, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[11])
```

## WFA vs Control


```{r DE_comp12, fig.height=8, fig.width=9}
DE_gene_by_lappy(current_contrast = list_with_names[12])
```

# Session Info
```{r session_info, echo = TRUE, eval=TRUE}
sessionInfo()
```

